---
title: "Introduction to the refdb package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the refdb package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 5,
  fig.width = 7
)
```


Developing and maintaining a DNA reference database can be difficult. The `refdb` package has been developed to assist you during the different steps of the process. This vignette will present some of the important functions of the package and help you to get started with it.


## Installing and loading the package

The refdb package is available on CRAN which means that you can easily install the latest stable version using the RStudio IDE or with the function `install.packages`. In this tutorial we will also be using functions from several packages of the tidyverse. Therefore, if the `tidyverse` package is not already installed on your computer, you are encouraged to install it now.


Once everything is ready, we can start by loading the packages we need:

```{r setup}
library(refdb)
library(tidyverse)
```

### Reading data

#### Reading data from a file


In refdb a reference database is a simple dataframe. This means that any functions able to read external data and return a dataframe in your R environment can be used to import a reference library. Here, for simplicity, we will assume the reference database is stored as a CSV on the disk. You can download this file here. Then, we can read the data from their location on your computer with:

```{r, include=FALSE}
lib <- read_csv(system.file("extdata", "ephem.csv", package = "refdb"))
```
```{r, eval=FALSE}
lib <- read_csv("my_path/ephem.csv")
```


Note that we use the function `read_csv` from tidyverse which returns a tibble. A tibble is a dataframe with some extra refinements. There is no issue with using tibbles with refdb functions (this is actually encouraged).


#### Setting fields

Now we have a dataframe (a tibble) in our environment. However, the functions of the refdb package can do nothing from it because they need to know what is the actual content stored in each column. You have to provide this information, and this can be done using the function `refdb_set_fields`.

```{r}
lib <- refdb_set_fields(lib,
                        taxonomy = c(family = "family_name",
                                     genus = "genus_name",
                                     species = "species_name"),
                        sequence = "DNA_seq",
                        marker = "marker")
```

Here we explicitly set the taxonomy, sequence and marker fields by indicating which columns of the dataframe correspond to each of those fields.

If you print the object `lib` now, you will see that the nucleotide symbols in the columns DNA_seq are colored. This is because this column has been identified as containing DNA sequences. A simple way to see which fields have been associated with which column is simply to use the function `refdb_get_fields`.

```{r}
refdb_get_fields(lib)
```


### Cleaning taxonomic names and DNA sequences

Now that the data are loaded and the fields have been set, we can start to work with the reference database. The first thing we will look at, is how to clean taxonomic names and DNA sequences. Cleaning is different than filtering (that we will see in the next section) in that the objective here is not to remove entries but to improve them.


Both sequences and taxonomic names can be affected by countless problems that would be impossible to list here. However, the refdb package provides functions to deal with many of the common cases. Let's review some of them.

#### Missing taxonomic information

Let's take a look at the taxonomy of the 91st record:
```{r}
lib [91, 1:4]
```
Here, the specimen has been identified at the genus level and the species is reported using the notation "sp.". However to manage a database it is not really practical to use such notation. How would you report the family of an organism identified at the class level? A general and efficient way to code for a missing value is simply to use the built-in `NA` indicator.

The function `refdb_clean_tax_NA` can be used to identify taxonomic levels for which there are no information and change them to `NA`.

```{r}
lib_na <- refdb_clean_tax_NA(lib)
lib_na [91, 1:4]
```

#### Extra information appended to taxonomic names

It is common to find taxonomic names to which various information has been added. See the 32nd row of our database:

```{r}
lib [32, 1:4]
```

The species name is tagged "B BG" which probably meant something to the person who recorded this entry but is not useful to us and may even be problematic. The function `refdb_clean_tax_remove_extra` uses some simple rules to detect these information and remove them.

```{r}
lib_na_extra <- refdb_clean_tax_remove_extra(lib_na)
lib_na_extra [32, 1:4]
```




#### A note on false positive and negative

The cleaning functions of refdb uses several rules and heuristics, some of them can produce false positive and negative. You should always be careful and if you find cases where a function does not what it is expected to do, please report [here](https://github.com/fkeck/refdb/issues).



#### Cleaning DNA sequences



#### Other cleaning functions

We have only seen a brief overview of the functions available for cleaning up the sequences and their associated taxonomy. It would be tedious to list all the functions and their options here. Instead, the table below summarises the functions currently available in refdb.

TABLE

### Filtering records

Filtering is the step that usually follows cleaning and whose objective is to identify and remove records (rows) which do not satisfy quality criteria (those criteria and associated thresholds are of course specific to each project).

For example we may want to exclude very short sequences whose length is likely shorter than the targeted genetic marker. Let's take a look at the distribution of sequence lengths in the database:

```{r, fig.width=4.5, fig.height=3.5}
refdb_plot_seqlen_hist(lib)
```

To exclude all the sequences shorter than 500 nucleotides we use the function `refdb_filter_seq_length`.
```{r}
lib_long <- refdb_filter_seq_length(lib, min_len = 500)

nrow(lib)
nrow(lib_long)
```
The function has removed `r nrow(lib) - nrow(lib_long)` records.

Another, more sophisticated filtering function is `refdb_filter_seq_primer()` which can find sequences which do not contain a motif (typically a primer sequence) and remove them from the reference database

For example, a famous forward primer for metazoan is the mlCOIintF (Leray 2013). I we want to keep only the sequences where this primer is found we can use:

```{r}
refdb_filter_seq_primer(lib, primer_forward = "GGWACWGGWTGAACWGTWTAYCCYCC")
```

You can see that only 76 sequences have passed the filter. By default the tolerance of error is 10%. We can change this value using the `max_error_forward` argument.

```{r}
refdb_filter_seq_primer(lib, primer_forward = "GGWACWGGWTGAACWGTWTAYCCYCC", max_error_forward = 0.2)
```

By increasing the maximum error rate to 20%, we now retain 99 sequences.


### Visualizing the reference library

Now let's take a tour of the functions you can use to produce graphical representation of your reference database. The fact that refdb stores reference database as dataframes makes it really easy to produce plots (e.g. with the tidyverse and `ggplot2`). For example, we can make a barplot showing the distribution of the families like this:

```{r}
lib %>% 
  group_by(family_name) %>% 
  count() %>% 
  ggplot(aes(fct_reorder(family_name, n, .desc = TRUE), n)) +
  geom_col() +
  xlab("Family") +
  ylab("Number of records") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

Additionally, refdb provides ready-to use functions to produce more sophisticated plots. For example to represent the taxonomic coverage of your reference library across multiple levels with a tree map.

```{r, message=FALSE, warning=FALSE}
refdb_plot_tax_treemap(lib)
```

Alternatively you can represent this information with a taxonomic tree. These functions have several parameters to control what is represented in the plot.

```{r, message=FALSE, warning=FALSE}
refdb_plot_tax_tree(lib)
```

We conclude our overview of the graphical features of refdb with the refdb_plot_map function which can be used to generate interactive maps from the data. Of course this requires the sequences to be geographically located. As we have not yet associated columns to the latitude and longitude fields, we have to do this beforehand.

```{r, message=FALSE, warning=FALSE}
lib <- refdb_set_fields(lib, latitude = "lat", longitude = "lon")
refdb_plot_map(lib)
```

### Creating a report

A report is a simple and rapid way to get an overview of the current state of your reference library and to identify some possible issues. You can compile a report using the function `refdb_check_report`.

```{r, eval=FALSE}
refdb_check_report(lib)
```


The result (not shown here) is an interactive HTML report which can be opened in any recent web browser. It contains some statistics and plots, and the results of functions `refdb_check_tax_typo`, `refdb_check_tax_conflict` and `refdb_check_seq_conflict`. These functions can be used to identify possible spelling errors in taxonomic names, conflicts in the taxonomic tree and lack of genetic resolution in sequences, respectively.


### Exporting the data

Once satisfied with the results, they can be exported for further use. Since the reference database is a dataframe, it is possible to use any export functions (from R or from other packages) that support dataframes. Here we show how to export to CSV using the function `write_csv` from the `readr` package.

```{r, eval=FALSE}
write_csv(lib, "my_reference_library.csv")
```

This will write the actual content of `lib` but ignore the fields which have been previously set. We can save them in a separate YAML file using:

```{r, eval=FALSE}
refdb_write_fields(lib, "my_reference_fields.yml")
```

Then, the next time you will open this reference database, it will be easy to set the fields again using the argument `config_yaml` of the function `refdb_set_fields`:

```{r, eval=FALSE}
lib <- read_csv("my_reference_library.csv")
lib <- refdb_set_fields(lib, config_yaml = "my_reference_fields.yml")

refdb_get_fields(lib)
```








